---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `ieegio`: IO toolbox for Intracranial Electroencephalography

<!-- badges: start -->
[![R-CMD-check](https://github.com/dipterix/ieegio/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dipterix/ieegio/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of `ieegio` is to provide integrated toolbox for common file formats used in intracranial Electroencephalography (iEEG) and deep-brain stimulation (DBS). Currently the package is under active development The following formats have been (or will be) supported:

**Electrophysiology**

| Name                                        |            Common Extensions            |        Read        |       Write        |
|:--------------------------------------------|:---------------------------------------:|:------------------:|:------------------:|
| EDF(+)/BDF(+) (European Data Format)        | `*.edf`                                 | :white_check_mark: |                    |
| BrainVision                                 | `*.vhdr`, `*.vmrk`, `*.eeg`, `*.dat`    | :white_check_mark: |                    |
| BCI2000                                     | `*.dat`                                 | :white_check_mark: |                    |
| NEV/NSx (BlackRock neural-event/signal)     | `*.nev`, `*.ns1`, `*.ns2`, ..., `*.ns6` | :white_check_mark: |                    |
| NWB (Neurodata Without Borders)             | `*.nwb`                                 | :white_check_mark:$^{*}$ | :white_check_mark:$^{*}$ |
| MEF                                         | `*.mef`                                 |                    |                    |

* `NWB` format only has low-level support. Please see examples with `help(read_nwb)`

**Imaging**

| Name                                        |            Common Extensions            |        Read        |       Write        |
|:--------------------------------------------|:---------------------------------------:|:------------------:|:------------------:|
| NIfTI (Neuroimaging Informatics Technology Initiative) | `*.nii[.gz]`                 | :white_check_mark: | :white_check_mark: |
| MGH/MGZ (Massachusetts General Hospital format) | `*.mgh`, `*.mgz`                    | :white_check_mark: | :white_check_mark: |
| GIfTI (Geometry format under the Neuroimaging Informatics Technology Initiative) | `*.gii[.gz]` | :white_check_mark: |  |
| FreeSurfer surface                          | e.g. `*h.pial`, `*h.white`              | :white_check_mark: |  |
| FreeSurfer node value                       | e.g. `*h.curv`, `*h.sulc`               | :white_check_mark: |  |
| Surface annotation                          | `*.annot`                               | :white_check_mark: |  |

Please feel free to make a wish-list by posting an issue in this repository.

**Miscellaneous**

| Name        | Common Extensions |        Read        |       Write        |
|:------------|:-----------------:|:------------------:|:------------------:|
| Matlab      | `*.mat`           | :white_check_mark: | :white_check_mark: |
| HDF5        | `*.h5`            | :white_check_mark: | :white_check_mark: |
| YAML        | `*.yml`, `*.yaml` | :white_check_mark: | :white_check_mark: |
| JSON        | `*.json`          | :white_check_mark: | :white_check_mark: |
| INI         | `*.ini`           | :white_check_mark: |                    |
| FST         | `*.fst`           | :white_check_mark: | :white_check_mark: |


## Installation

You can install the development version of `ieegio` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("dipterix/ieegio")
```

`ieegio` provides additional sample data. To download them:


``` r
# list all the sample data
sample_names <- ieegio::ieegio_sample_data()

lapply(sample_names, ieegio::ieegio_sample_data)
```

## Example 1: Read Formatted `ephys` Data

This is a basic example which reads sample `EDF` data

```{r}
library(ieegio)
edf_path <- ieegio_sample_data("edfPlusD.edf")
edf <- read_edf(edf_path, verbose = FALSE)
edf
```

You can check header, channel table, and annotations

```{r}
header <- edf$get_header()
chan_tbl <- edf$get_channel_table()
annot <- edf$get_annotations()
```

You can also query a channel by:

```{r}
# get Channel 1
channel <- edf$get_channel(1)
channel
```

Let's plot the channel

```{r}
plot(
  x = channel$time, y = channel$value,
  xlab = "Time", ylab = channel$info$Unit,
  main = channel$info$Label,
  type = "l"
)
```

Most formats (`EDF`, `BrainVision`, `BlackRock`, `BCI2000`) provide similar syntax, with an exception `NWB` since `NWB` is essentially folder-based format that is enclosed within an `HDF5` file. Check documentations and examples in `help(read_nwb)` for more details.

## Example 2: Imaging Data

Each volume data (`NIfTI`, `MGH`, ...) contains `header`, `data`, and `transforms` list.

```{r}
library(ieegio)
nii_path <- ieegio_sample_data("brain.demosubject.nii.gz")
nii <- io_read_nii(nii_path)
nii
```

The header object is the low-level handler

```{r}
nii$header
```

The transforms contain transforms from volume (column, row, slice) index to other coordinate systems. The most commonly used one is `vox2ras`, which is a `4x4` matrix mapping the voxels to scanner (usually `T1-weighted`) `RAS` (right-anterior-superior) system.

```{r}
nii$transforms$vox2ras
```

The data object contains the volume image

```{r}
image(nii$data[, 128, ], asp = 1, axes = FALSE, 
      col = colorRampPalette(c("black", "white"))(255))
```
